{\bf \huge Wikiplia}:\\ The Free Programming Language\\ That Anyone Can Edit

### 2007-03-16 trunk/wikiplia/paper/wikiplia.tex
working

\section{Reflections on booting boots} \label{sec:bootstrap}

>>>>

\section{Reflections on strapping straps and booting boots} \label{sec:bootstrap}

The hallmark of Free software is the GNU General Public License. It is
a hereditary license that requires that (a) source code be distributed
with the program and (b) modified versions of the program also be
licensed under the GPL. The intention is that anyone receiving the
software can exercise Freedom~$\alpha$ by understanding the source
code and modifying it to suit his needs. Clearly any source code will
not do: an obfuscated\wcite{Obfuscated code} version of the source
code cannot be easily understood or modified, even though it is
technically ``source code.'' The GPL therefore defines source code as
the ``preferred form'' for ``making modifications.''

Even source code in the preferred form might not be enough to achieve
Freedom~$\alpha$, however. For instance, the program might be written
in a mysterious programming language that only the author understands,
and that programming language might only . It is therefore reasonable
to construe the ``preferred form'' of the original software to include
the implementation of the programming language that the software is
written in. Because the programmer might need to fix bugs or extend
the programming language implementation in order to modify the
original program, he also needs the source code for that language as
well. This code must be written in some language as well, so the
process continues. It can end when one of the programming languages is
generally well known enough that there are no practical barriers to
understanding it or finding an implementation (examples would include
C\wcite{C (programming language) and ALGOL 58\wcite{ALGOL 58}}, or so
simple that the implementation is essentially non-existent ({\em e.g.}
an assembler implemented directly in machine code).

Another way for this process to terminate is for a programming
language to be implemented in itself. This is known as a
``bootstrapping compiler.'' A natural social tendency causes this to
be very common: language implementors are more likely to enjoy the
language they are implementing, and therefore more likely to choose it
to implement the language. But when this process terminates this way,
the reader might be left with a suspicious sense that nothing has
actually been achieved.\z\footnote{In the case of a LOGO interpreter
implemented in LOGO, we could say that this is then ``turtles all the
way down.''} Specifically: What freedom-fulfilling use is the source
code to an implementation of a mysterious programming language, if
that source code is itself in the same mysterious programming
language?

Let us concentrate on a more concrete example. The GNU C
compiler\z\wcite{GNU Compiler Collection} (licensed under the GPL) is
an implementation of the C langauge with some extensions specific to
the compiler. The GCC source code uses some of these extensions to HERE

% parallel development yields same source, different behavior.
% need globally atomic commits across all developers

All popular modern languages are defined via a definitional
interpreter\wcite{JavaScript}\wcite{Objective Caml}\wcite{Perl} with
accompanying O'Reilly ``animal'' book.
